Utility of Diffusion Models for Predicting and Diagnosing the Spread of Conservation Behaviors

### 2021-10-16 DOI_Review/Manuscript.rmd
Tons of updates on DOI review and HCRI proj.

---
title: Utility of Diffusion Models for Predicting and Diagnosing the Spread of Conservation Behaviors
author:
- affiliation: Boise State University - Human Environment Systems
  name: Matt Clark
- affiliation: Boise State University- Human Environment Systems
  name: Vicken Hillis
- affiliation: Max Planck Institute for Evolutionary Anthropology
  name: Jeffrey Andrews

date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Community-based conservation, diffusion of innovations, cultural evolution, complex systems
bibliography: [./DOI_Conservation_Review.bib]
abstract: Conservation researchers are increasingly identifying diffusion theory as useful for understanding the spread of community-based conservation initiatives. This work has nevertheless been largely ambiguous as to the specific opportunities and shortcomings for diffusion theory in conservation research, and how it might be practically applied. Here, we employ agent-based simulations to confirm that process-based diffusion models reasonably predict the long-term patterns in adoption of conservation behaviors even under tightly coupled behavior-environment feedbacks. We caution however, that while diffusion theory may predict the pattern with which a conservation intervention will spread through a population, it gives little insight as to the underlying transmission mechanism. We show that the transmission mechanism employed by a population can alter the final proportion of individuals that adopt an intervention, and argue that conservation researchers and practitioners can leverage theory from social learning and cultural evolution to better design and diagnose projects. Lastly, we emphasize that developing and fitting process models in statistical frameworks traditionally reserved for regression alone, may pose a productive new avenue for conservation science, particularly in our understanding of conservation as a complex system.

@@

> "Diffusion study...is something like the use of radioactive tracers in studying the process of plant   growth: it helps illuminate the process."
>
> `r tufte::quote_footer('--- Everett M. Rogers, Diffusion of Innovations, 2003')`

@@

While some debate still exists and exact methods vary, it is largely accepted that community-based conservation interventions are a viable method for seeding self-spreading, -sustaining, and ethical protections on natural resources [@turner_global_2007; @berkes_community-based_2007; @adams_biodiversity_2004]. A major question in implementing these projects is if a specific intervention will indeed catch-on in a population and if so, how quickly [@radeloff_hot_2013]? Predicting the final proportion of a population that will adopt an intervention is critical, as many interventions require buy-in from nearly all community members to be effective [@mangubhai_not_2011; @visconti_protected_2019].

# Diffusion Theory
Inducing and/or speeding up the widespread adoption of some novel (to the individual) behavior is not an issue unique to community-based conservation; producers of cigarettes and snack foods, as well as public health professionals, have been grappling with these questions for quite some time. Diffusion theory recognizes that this process is primarily governed by the level of uncertainty an individual has about a novel behavior and their personal threshold for the amount of certainty required to make a decision [@rogers_diffusion_2010]. When an individual is introduced to a new behavior, they often have a high degree of uncertainty as to the risks and benefits of adopting and will wait to adopt until that uncertainty is reduced to a tolerable level. The primary way individuals reduce uncertainty about the outcome of adopting a particular behavior is through second-hand observation of early adopters. Therefore, the core of diffusion theory is the notion that individuals imitate those around them when making decisions about adopting a novel behavior or not, rather than weighing the pros and cons independently.

This assumption about human decision-making has been validated repeatedly in cultures around the world; individuals do not act as cost-benefit analysts, in which they adapt their behaviors based on the environment alone [@henrich_search_2001]. Our dependence on social learning to make decisions about behavioral adoption has profound and specific implications for patterns of behavioral adoption at the population level [@smaldino_adoption_2017]. We observe these patterns through diffusion curves which show the proportion of a population that has adopted a behavior as a function of time [@hoppitt_effect_2010; @rogers_diffusion_2010; @ryan_acceptance_1950].

Behavioral spread through imitation produces a sigmoidal “S” shaped diffusion curve, which is observed in the vast majority of empirical studies on behavioral adoption [@henrich_cultural_2001]. This “S” shaped curve is produced regardless of the social learning mechanism employed by a population and can even be observed when a portion of the population relies on individual learning [@henrich_cultural_2001]. Behavioral adoption thus exhibits equifinality, where many different underlying mechanisms may produce similar or even identical patterns over time [@barrett_equifinality_2019].

@@

Individuals tend to place more weight on imitation when making decisions in which the outcomes can not be well predicted independently [@perreault_bayesian_2012]. This phenomenon is particularly salient under variable, rapidly changing environments [@boyd_culture_1988; @richerson_not_2008]. It follows then, that the adoption of novel conservation behaviors by a community is likely to be achieved through social imitation (i.e. diffusion), especially given worsening climatic instability and regime changes resulting from widespread biodiversity loss.

Indeed, the sigmoidal pattern of behavioral adoption characteristic to diffusion has been observed in over 80% of case studies on community-based conservation interventions worldwide [@mascia_when_2018; @mills_how_2019]. Conversely, in contexts where communication and thus imitation are limited, such as across island chains, the adoption of community based conservation interventions is considerably attenuated [@pietri_information_2009-1]. With the recognition that the patterns of uptake for community-based conservation practices are directly linked to behavioral imitation, researchers have begun suggesting diffusion theory as a lens for understanding and implementing individual interventions [@abernethy_two_2014; @mbaru_key_2017; @niemiec_efficacy-based_2021].

@@

These studies are correct in their conclusion that diffusion theory can be productively applied to the spread of community-based conservation interventions, but ambiguous in exactly how a practitioner or researcher might practically apply this insight. The aim of this review is to explicitly demonstrate how diffusion theory can be used to generate predictions about the patterns of uptake of conservation behaviors, and what the limitations of this theory are for complex adaptive systems like community-based conservation.

@@

The sigmoidal pattern of behavioral spread over time mirrors that of disease transmission [@smaldino_coupled_2020]. This parallel implies that the well-established mathematics used to describe patterns of disease transmission can be applied to fitting models of behavioral uptake [@cavalli-sforza_cultural_1981; @strang_adding_1991]. Specifically, the canonical susceptible-infectious-recovered (SIR) model and SIRS variation where individuals are added back to the pool of susceptibles after recovery (eq 1), have been demonstrated to be well suited to fitting the diffusion process [@famil-rohani_development_2019; @smaldino_coupled_2020]. In this instantiation, $\beta$ represents the rate of infection in the population, $\gamma$ represents the rate of recovery, and $\chi$ is the rate that recovered individuals are added back into the pool of susceptibles.

\begin{align*}
&\frac{dS}{dt} = - \beta * \text{S}_{t} * \text{I}_{t} + \chi * \text{R}_{t} \\
&\frac{dI}{dt} = \beta * \text{S}_{t} * \text{I}_{t} - \gamma * \text{I}_{t} \\
&\frac{dR}{dt} = \gamma* \text{I}_{t} - \chi * \text{R}_{t}
\end{align*}

When the natural history characteristics of some pathogen is known, the SI, SIR, SIS, &  SIRS suite of models reliably give accurate, deterministic predictions about the number of individuals in a population who will become infected and for how long a disease will persist within a population [@huppert_mathematical_2013]. For conservation interventions however, there are no natural history features of a specific intervention, such as the rate or duration of adoption, which can be known ahead of time. We can however, use Bayesian inference to probabilistically estimate these parameters from data on behavioral uptake over time [@margossian_review_2019; @margossian_differential_2017;  @carpenter_predator-prey_2018]. Below, we use a minimal agent-based model of conservation adoption contingent on behavior-environment feedbacks to simulate data on the long-term adoption dynamics of a community-based conservation intervention. Then, using just the first 30% of the data, we estimate the rate of intervention spread, rate of dropout, and time after dropout before agents are available to adopt again. We use these estimated parameters to predict the overall patterns of adoption over all time steps and compare our observed to predicted values.

# Agent-based simulation
We developed a minimal agent-based simulation which we use to generate data on conservation adoption within a population. We use these data to test the performance of the SIRS diffusion model in predicting patterns of conservation behavior adoption under both unbiased and success-biased social learning. Success-biased social learning is contingent on behavior-environment feedbacks. Our simulated system emulates the common performance payments or payments for ecosystem services approach to conservation [@ferraro_future_2011]. We establish a degraded landscape/seascape (from herein referred to as landscape) where all simulated individuals are dependent on harvesting some renewable, mobile resource. In the first timestep, we introduce a conservation intervention which delineates 20% of the landscape as a conservation area and individuals are rewarded for abstaining from harvesting there. In each subsequent timestep, agents either choose to enroll in the conservation program, receive their reward, and continue to harvest just from the remaining 80% of the unprotected landscape, or they choose to harvest from the entire landscape and forgo the reward. After each harvest period, agents are randomly paired, and employ either unbiased or success-biased learning to decide whether or not to adopt the conservation behavior [@baldini_success-biased_2012; @baldini_two_2013, @barrett_equifinality_2019]. The protected and unprotected resource pools are reduced according to agents’ harvesting behavior, regenerate, and move from one area to the other if the carrying capacity is reached in any respective area.

@@

A visual description of this minimal model is shown in figure 1, and the full ODD+ model description can be found in the supplemental material (S1). Interested readers can find an interactive web application which allows users to run this simulation given a wide variety of starting conditions at this URL [https://matthewclark.shinyapps.io/ConservationAdoptionSimulation/](https://matthewclark.shinyapps.io/ConservationAdoptionSimulation/). As this simulation is not the emphasis of this study, we show just two runs below, one in which. The starting conditions of these runs are identical except for the social learning mechanism employed by the population. As the unbiased learning mechanism is independent of the resource levels, the adoption curve follows the traditional “S” pattern and nears fixation in the long-term. Under environmentally mediated success-biased learning, the adoption pattern spikes initially, but drops off dramatically as the resources in the unprotected landscape are reduced following increased harvest pressure there as a result of the conservation intervention. The full output of this later simulation is shown below in figure 2. Additional simulation runs can be found in the supplemental material (S2) or by using the interactive web application linked above.

@@

$$ \text{Y}_{t} \sim Binomial(\text{N}, \text{I}_{t}) $$
\begin{align*}
&\frac{dS}{dt} = - \beta * \text{S}_{t} * \text{I}_{t} + \chi * \text{R}_{t} \\
&\frac{dI}{dt} = \beta * \text{S}_{t} * \text{I}_{t} - \gamma * \text{I}_{t} \\
&\frac{dR}{dt} = \gamma* \text{I}_{t} - \chi * \text{R}_{t}
\end{align*}

@@

For both learning mechanisms the SIRS model exhibited convergence for all chains and resulted in an rhat of 1 for all estimated parameters (fig 3). As the parameters $\beta$, $\gamma$, and $\chi$ are directly interpretable as the rates of adoption, dropout, and willingness to retry a previously abandoned intervention, accurate estimation of these values alone, especially in the early stages of an intervention, is a practical and straightforward application of diffusion theory to community-based conservation interventions.

@@

We can also use these parameter values, estimated from the first 30% of the time series, to predict the long-term patterns of adoption for community-based conservation initiatives. In figures 4 & 5, the translucent black lines show 1000 draws from the posterior predicting the proportion of the population who will be enrolled in the intervention at every time step. Grey and blue points show the initial 30% of the data which was used to estimate the three parameters of interest and the remaining 70% of data respectively. These figures show that the SIRS model accurately captures and predicts the pattern of behavioral adoption for both unbiased learning and environmentally mediated success-biased learning. R and STAN code for these models can be found in the supplemental material (S4)

@@

Figures 4 & 5 also demonstrate that different learning mechanisms, while both producing a sigmoidal adoption pattern, can result in different overall rates in adoption of an intervention, even under identical starting conditions. In the simulation of a population employing unbiased learning, the population fixates at nearly 90% adoption of the intervention behavior. For success-biased learning on the other hand, resource scarcity in the unprotected area caused the population to adopt the intervention at approximately 60% in the long-term.

@@

As demonstrated above, the standard SIRS diffusion model can reliably capture the long-term patterns of behavioral adoption for community-based conservation interventions, even when adoption is contingent on feedbacks with ecological processes. This model, while accurately capturing the pattern of spread, is unable to provide insight as to the mechanism of spread or causality in the system. As shown above and in existing literature, the social transmission mechanism, or mix of mechanisms, employed by a population can dramatically affect the overall rates of adoption for a given behavior [@smaldino_adoption_2017; @andrews_cultural_2018; @tam_measuring_2021]. Therefore, while we agree that diffusion theory can be productively applied to community-based conservation, we point to theories of social learning and cultural evolution as a better alternative for diagnosing and planning interventions.

Understanding the causal mechanisms behind the patterns of behavioral transmission is necessary to better delineate and predict how conservation will affect both people and the environment [@cheng_strengthen_2020]. Researchers are increasingly recognizing that models of human-caused environmental change must explicitly incorporate dynamic social processes [@turner_land-useland-cover_2021]. Specifically, in modeling social-ecological systems, either using analytical approaches or simulation, processes should be based on realistic human cognitions which allow actors to modify their behaviors in response to new information [@meyfroidt_environmental_2013]. Modeling human behavior in the context of community-based conservation is thus going to require a complex dynamic systems approach in which behaviors and the biophysical environment coevolve [@schill_more_2019].

We argue that when prediction is the goal, analytical or statistical approaches like the SIRS model presented above are always preferable to simulation based approaches [@epstein_why_2008; @tredennick_practical_2021]. The practice of applying statistical frameworks of parameter estimation to analytical process models, and in particular to ordinary differential equations, is still relatively new and has been embraced mostly by ecology [@hefley_when_2017]. Extending the SIRS diffusion model to include a term which represents changes in the environment over time and the effect of resource levels on adoption rates may more accurately describe the system of adoption in community-based conservation interventions. An example of such an extension is presented below where we include another equation for the change in a resource pool over time $\frac{dE}{dt}$. We also must include three additional parameters: $\phi$. $\rho$, and $\theta$, which represent the effect of resource level on enrollment (infection), the effect of harvest on the resource level, and the regeneration rate of the resource respectively.

\begin{align*}
&\frac{dS}{dt} = - \beta * \text{S}_{t} * \text{I}_{t} + \chi * \text{R}_{t} \\
&\frac{dI}{dt} = \beta * \text{S}_{t} * \text{I}_{t} - \gamma * \text{I}_{t} - \phi * \text{E}_{t} \\
&\frac{dR}{dt} = \gamma * \text{I}_{t} - \chi * \text{R}_{t} + \phi * \text{E}_{t} \\
&\frac{dE}{dt} =  - \rho * (\text{S}_{t} + \text{R}_{t}) + \theta*\text{E}_{t}
\end{align*}

While process models may be preferable for predicting adoption outcomes of conservation interventions, they might not always be possible. In complex systems, these models can quickly require many parameters in order to accurately account for all moving parts and thus are prone to issues of identifiability in parameter estimation [@laubmeier_ecological_2020]. Further, complex social-ecological systems have properties which are inherently stochastic and may be impossible to model analytically. For example, coastal fisheries have been shown to approach mathematical chaos in the number and proportions of species present [@wilson_chaos_1994].

In very complex cases such as these, we argue that agent-based simulations are the preferred approach as they allow for tipping points and emergence, two key features of the complex systems which can not be accounted for with process models such the SIRS or SIRS extension models presented above [@reyers_social-ecological_2018; @railsback_agent-based_2012; @wilensky_introduction_2015]. Coupled human and natural systems commonly exhibit dynamics with thresholds, resilience, and heterogeneity not present in social or natural systems in isolation and therefore are difficult to model analytically  [@liu_complexity_2007].

>>>>

---
title: Utility of Diffusion Models for Predicting and Diagnosing the Spread of Conservation Behaviors
author:
- affiliation: Boise State University - Human Environment Systems
  name: Matt Clark
- affiliation: Boise State University- Human Environment Systems
  name: Vicken Hillis

date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Community-based conservation, diffusion of innovations, cultural evolution, complex systems
bibliography: [./DOI_Conservation_Review.bib]
abstract: A common goal of conservation practitioners is to design interventions that benefit both people and the environment, and as a result are adopted robustly by local communities. Diffusion theory provides generalized heuristics and models for how learned behaviors spread within a population and has been increasingly identified by researchers as useful for understanding conservation adoption. Current prescriptions have nevertheless been largely ambiguous as to the practical applications of diffusion theory to conservation work, and in particular, in coupled human-natural systems. Here, we demonstrate the capacity for a process-based diffusion model to reasonably predict the long-term rates of conservation adoption under tightly coupled behavior-environment feedbacks. Data on conservation adoption are generated through an agent-based simulation which highlights our dynamics of interest. We caution however, that while diffusion models may predict the pattern with which a conservation intervention will spread through a population, they give little diagnostic insight as to why. We show that the social learning mechanism employed by a population can alter the final proportion of individuals that adopt an intervention, and argue that conservation researchers and practitioners can leverage theory from social learning and cultural evolution to better design and diagnose projects. Lastly, we emphasize that developing and fitting process models in statistical frameworks traditionally reserved for linear regression alone, may pose a productive new avenue for conservation science, particularly in concert with simulation techniques and insights from complex systems.

@@

> "Diffusion study...is something like the use of radioactive tracers in studying the process of plant   growth: it helps illuminate the process."
>
> `r tufte::quote_footer('--- Everett M. Rogers, Diffusion of Innovations, 2010')`

@@

While some debate still exists and exact methods vary, it is largely accepted that community-based conservation interventions are a viable method for seeding self-spreading, -sustaining, and ethical protections on natural resources [@turner_global_2007; @berkes_community-based_2007; @adams_biodiversity_2004]. A major question in implementing these projects is if a specific intervention will indeed catch on in a population and if so, how quickly [@radeloff_hot_2013]? Predicting the final proportion of a population that will adopt an intervention is critical, as many interventions require buy-in from nearly all community members to be effective [@mangubhai_not_2011; @visconti_protected_2019].

# Diffusion Theory
Promoting the widespread adoption of some novel (to the individual) behavior is not an issue unique to community-based conservation; producers of cigarettes and snack foods, as well as public health professionals, have been grappling with these questions for quite some time. Diffusion theory recognizes that this process is primarily governed by the level of uncertainty an individual has about a novel behavior and their personal threshold for the amount of certainty required to make a decision [@rogers_diffusion_2010]. When an individual is introduced to a new behavior, they often have a high degree of uncertainty as to the risks and benefits of adopting and will wait to adopt until that uncertainty is reduced to a tolerable level. The primary way individuals reduce uncertainty about the outcome of adopting a particular behavior is through second-hand observation of early adopters. Therefore, the core of diffusion theory is the notion that individuals imitate those around them when making decisions about adopting a novel behavior or not, rather than weighing the pros and cons independently.

This assumption about human decision-making has been validated repeatedly in cultures around the world; individuals do not act as cost-benefit analysts, in which they adapt their behaviors based on the environment alone [@henrich_search_2001]. Our dependence on social learning to make decisions has profound and specific implications for patterns of behavioral adoption at the population level [@smaldino_adoption_2017]. We observe these patterns through diffusion curves which show the proportion of a population that has adopted a behavior as a function of time [@hoppitt_effect_2010; @rogers_diffusion_2010; @ryan_acceptance_1950].

Behavioral spread through imitation produces a sigmoidal “S” shaped diffusion curve, which is observed in the vast majority of empirical studies on behavioral adoption [@henrich_cultural_2001]. This “S” shaped curve is produced regardless of the social learning mechanism employed by a population and can even be observed when a portion of the population relies on individual learning [@henrich_cultural_2001]. Behavioral adoption thus exhibits equifinality, where many different underlying mechanisms may produce similar or even identical patterns over time [@barrett_equifinality_2019]. Critically, this makes it virtually impossible to infer the learning mechanism(s) used by a population from the pattern of adoption over time alone.

@@

Individuals tend to place more weight on imitation when making decisions in which the outcomes can not be easily predicted independently [@perreault_bayesian_2012]. This has been explicitly demonstrated under variable, rapidly changing environments [@boyd_culture_1988; @richerson_not_2008]. It follows then, that the adoption of novel conservation behaviors by a community is likely to be achieved through social imitation (i.e. diffusion), especially given worsening climatic instability and regime changes resulting from widespread biodiversity loss.

Indeed, the sigmoidal pattern of behavioral adoption characteristic to diffusion has been observed in over 80% of case studies on community-based conservation interventions worldwide [@mascia_when_2018; @mills_how_2019]. Conversely, in contexts where communication and thus imitation are limited, such as across island chains, the adoption of community-based conservation interventions is considerably attenuated [@pietri_information_2009-1]. With the recognition that the patterns of uptake for community-based conservation practices are directly linked to behavioral imitation, researchers have begun suggesting diffusion theory as a lens for understanding and implementing individual interventions [@abernethy_two_2014; @mbaru_key_2017; @niemiec_efficacy-based_2021].

@@

These studies are compelling in their conclusion that diffusion theory can be productively applied to the spread of community-based conservation interventions, but ambiguous in exactly how someone might practically apply this insight. We envision a practitioner who is monitoring the weekly, monthly, or even yearly uptake rates of an intervention and is hoping to predict if and when it will “take off” in the population, what the final proportion of adopters will be, and/or what they can do to promote its spread. The aim of this review is to explicitly demonstrate how diffusion theory can be used to generate predictions about the patterns of uptake of conservation behaviors given realistic data, and the limitations of this theory as a diagnostic tool and for application to complex, coupled human-natural systems.

@@

The sigmoidal pattern of behavioral spread over time mirrors that of disease transmission [@smaldino_coupled_2020]. This parallel implies that the well-established mathematics used to describe patterns of disease transmission can be applied to fitting models of behavioral spread [@cavalli-sforza_cultural_1981; @strang_adding_1991]. Specifically, the canonical susceptible-infectious-recovered (SIR) model and SIRS variation where individuals are added back to the pool of susceptibles after recovery (eq 1), have been demonstrated to be well suited to fitting the diffusion process [@famil-rohani_development_2019; @smaldino_coupled_2020]. These compartmental differential equations assign every member of a population to a category (S, I, or R), depending on the proportion of individuals in the other categories and parameter values for the natural history characteristics of a given disease [@tolles_modeling_2020]. In this instantiation, the natural history parameter $\beta$ represents the rate of infection in the population, $\gamma$ represents the rate of recovery, and $\chi$ is the rate that recovered individuals are added back into the pool of susceptibles.

\begin{align}
&\frac{dS}{dt} = - \beta \cdot \text{S}_{t} \cdot \text{I}_{t} + \chi \cdot \text{R}_{t} \nonumber\\
&\frac{dI}{dt} = \beta \cdot \text{S}_{t} \cdot \text{I}_{t} - \gamma \cdot \text{I}_{t} \\
&\frac{dR}{dt} = \gamma \cdot \text{I}_{t} - \chi \cdot \text{R}_{t} \nonumber
\end{align}

When the natural history characteristics of some pathogen is known, the SI, SIR, SIS, & SIRS suite of models reliably give accurate, deterministic predictions about the number of individuals in a population who will become infected and for how long a disease will persist within a population [@huppert_mathematical_2013]. For conservation interventions however, there are no natural history features of a specific intervention, such as the rate or duration of adoption, which can be known ahead of time. We can however, use Bayesian inference to probabilistically estimate these parameters from data on behavioral uptake over time [@margossian_review_2019; @margossian_differential_2017; @carpenter_predator-prey_2018].

Below, we develop a minimal agent-based model to highlight our core dynamics of interest. In this abstraction, we simulate the spread in adoption of a conservation intervention from just a few initial individuals, to a stable proportion of the population. We use this simulation to produce realistic data on long-term adoption dynamics under tightly coupled behavior-environment feedbacks unique to community-based conservation interventions. Then, using just the first 30% of the data, we probabilistically estimate the rate of intervention spread, rate of dropout, and readoption using the SIRS model shown above. We use these estimated parameters to predict the long-term patterns of adoption for the remaining 70% of time steps and compare our predicted values to those observed in the simulation.

# Agent-based simulation
We developed a minimal agent-based simulation which we use to generate data on conservation adoption within a population. We use these data to test the performance of the SIRS diffusion model in predicting patterns of conservation behavior adoption under two social learning mechanisms. The first is unbiased copying, where agents adopt an alternative behavior with a probability proportional to its frequency in the population, independent of the benefit provided by adopting the behavior or not [@barrett_equifinality_2019]. The second mechanism is success-biased learning, under which agents disproportionately copy the behaviors of more successful agents [@baldini_two_2013]. This mechanism is formally demonstrated in equation 1, for two behaviors: $A$ & $B$, where the probability that an individual adopts behavior $A$ is a function of the mean pay-off, $\bar{P}$, of both behaviors at each time step and the strength of the bias, $\beta$ [@barrett_equifinality_2019].

\begin{equation}
 \text{Pr}(A|A,B)_{t+1} = \frac{\text{exp}\beta\bar{P}_{A,t}}{\text{exp}\beta\bar{P}_{A,t} + \text{exp}\beta\bar{P}_{B,t}}  
\end{equation}

Our simulated system emulates the common performance payments or payments for ecosystem services approach to conservation [@ferraro_future_2011]. We establish a degraded landscape/seascape (hereafter referred to as landscape) where all simulated individuals are dependent on harvesting some renewable, mobile resource. In the first time step, we introduce a conservation intervention which delineates 20% of the landscape as a conservation area and is initially adopted by just 2% of the population. In each subsequent time step, agents enrolled in the conservation program receive a reward, and continue to harvest just from the remaining 80% of the unconserved landscape. Agents not enrolled in the intervention harvest from the entire landscape and forgo the reward. After each harvest period, agents are randomly paired, and employ either unbiased or success-biased learning to decide whether or not to adopt the conservation behavior in the next time step [@baldini_success-biased_2012]. The conservation and unconserved resource pools are reduced according to agents’ harvesting behavior, regenerate, and move from one area to the other if the carrying capacity is reached in any respective area. As behavioral adoption under success-biased learning is dependent on harvest success, the success of each behavior is determined largely by the amount of available resources in each area, and the amount of resources in each area is a result of behavioral adoption, this learning strategy is contingent on behavior-environment feedbacks, unlike unbiased adoption which occurs independent of harvest success.

@@

A visual description of this minimal model is shown in figure 1, and the full ODD+ model description can be found in the supplemental material (S1). Interested readers can find an interactive web application which allows users to run this simulation given a wide variety of starting conditions at [https://matthewclark.shinyapps.io/ConservationAdoptionSimulation/](https://matthewclark.shinyapps.io/ConservationAdoptionSimulation/). As this simulation is not the emphasis of this study, we show just two runs below. The starting conditions of these runs are identical except for the social learning mechanism employed by the population. As the unbiased learning mechanism is independent of the resource levels, the adoption curve follows the traditional “S” pattern and nears fixation in the long-term. Under environmentally mediated success-biased learning, the adoption pattern spikes initially, but drops off dramatically as the resources in the unconserved landscape are reduced following increased harvest pressure there as a result of the conservation intervention. The full output of this latter simulation is shown below in figure 2.

@@

$$ \text{Y}_{t} \sim Binomial(\text{N}, \text{I}_{t}) $$
\begin{align}
&\frac{dS}{dt} = - \beta \cdot \text{S}_{t} \cdot \text{I}_{t} + \chi \cdot \text{R}_{t} \nonumber\\
&\frac{dI}{dt} = \beta \cdot \text{S}_{t} \cdot \text{I}_{t} - \gamma \cdot \text{I}_{t} \\
&\frac{dR}{dt} = \gamma \cdot \text{I}_{t} - \chi \cdot \text{R}_{t}\nonumber
\end{align}

@@

For both learning mechanisms the SIRS model exhibited convergence for all chains and resulted in an $\hat{R}$ of 1 for all estimated parameters (fig 3). As the parameters $\beta$, $\gamma$, and $\chi$ are directly interpretable as the rates of adoption, dropout, and willingness to retry a previously abandoned intervention, accurate estimation of these values alone, especially in the early stages of an intervention, is a practical and straightforward application of diffusion theory to community-based conservation interventions. For example, a practitioner might fit this model to real-world participation rates during the early weeks of an intervention in order to estimate whether they should focus more on increasing rates of uptake or retention going forward.

@@

We can also use these parameter values, estimated from the first 30% of the time series, to predict the long-term patterns of adoption for a community-based conservation initiative. In figures 4 & 5, the translucent black lines show 1000 draws from the posterior predicting the proportion of the population who will be enrolled in the intervention at every time step. Grey and blue points show the initial 30% of the data which was used to estimate the three parameters of interest and the remaining 70% of data respectively. These figures show that the SIRS model accurately captures and predicts the pattern of behavioral adoption for both unbiased learning and environmentally mediated success-biased learning. R and STAN code for these models can be found in the supplemental material (S4).

@@

Figures 4 & 5 also demonstrate that different learning mechanisms, while both producing a sigmoidal adoption pattern, can result in different overall rates in adoption of an intervention, even under identical starting conditions. In the simulation of a population employing unbiased learning, the population equilibrates at nearly 90% adoption of the intervention behavior. For success-biased learning on the other hand, resource scarcity in the unconserved area caused the population to adopt the intervention at approximately 60% in the long-term.

@@

As demonstrated above, the standard SIRS diffusion model can reliably capture the long-term patterns of behavioral adoption for community-based conservation interventions, even when adoption is contingent on feedbacks with ecological processes. This model, while accurately capturing the pattern of spread, is unable to provide insight as to the mechanism of spread or causality in the system. As shown above and in existing literature, the social transmission mechanism, or mix of mechanisms, employed by a population can dramatically affect the overall rates of adoption for a given behavior [@smaldino_adoption_2017; @andrews_cultural_2018; @tam_measuring_2021]. Therefore, while we agree that diffusion theory can be productively applied to community-based conservation, we point to theories of social learning and cultural evolution as a better alternative for diagnosing and planning interventions. For example, Tam et al. [-@tam_measuring_2021] and Kline et al. [-@kline_designing_2018] give practical examples of how to measure social learning strategies in the context of conservation, and tools (game/survey instrument, code, etc.) with which to collect and analyze the data.

Understanding the causal mechanisms behind the patterns of behavioral transmission is necessary to better delineate and predict how conservation will affect both people and the environment [@cheng_strengthen_2020]. Researchers are increasingly recognizing that models of human-caused environmental change must explicitly incorporate dynamic social processes based on realistic human cognitions which allow actors to modify their behaviors in response to new information [@meyfroidt_environmental_2013; @turner_land-useland-cover_2021]. Modeling human behavior in the context of community-based conservation is thus going to require a complex dynamic systems approach in which behaviors and the biophysical environment coevolve [@schill_more_2019].

The practice of applying statistical frameworks for parameter estimation to analytical process models, and in particular to ordinary differential equations, is still relatively new and has been embraced mostly in ecology [@hefley_when_2017; @laubmeier_ecological_2020]. This emerging methodology allows for the estimation of interesting and directly interpretable process parameters, not possible with linear regression. We believe this is an exciting new avenue for conservation science as the direct interpretability and applicability of process parameters lend themselves well to tight integration with applied research as is common in the field. Future research may be interested in extending the SIRS diffusion model to include a term which represents change in the environment over time and the effect of resource levels on adoption rates to more accurately reflect the social-ecological coupling. An example of such an extension is presented below, where we include another equation for the change in a resource pool over time $\frac{dE}{dt}$. We also must include three additional parameters: $\phi$, $\rho$, and $\theta$, which represent the effect of resource level on enrollment (infection), the effect of harvest on the resource level, and the regeneration rate of the resource respectively.

\begin{align}
&\frac{dS}{dt} = - \beta \cdot \text{S}_{t} \cdot \text{I}_{t} + \chi \cdot \text{R}_{t} \nonumber\\
&\frac{dI}{dt} = \beta \cdot \text{S}_{t} \cdot \text{I}_{t} - \gamma \cdot \text{I}_{t} - \phi \cdot \text{E}_{t} \\
&\frac{dR}{dt} = \gamma \cdot \text{I}_{t} - \chi \cdot \text{R}_{t} + \phi \cdot \text{E}_{t} \nonumber\\
&\frac{dE}{dt} = - \rho \cdot (\text{S}_{t} + \text{R}_{t}) + \theta \cdot \text{E}_{t}\nonumber
\end{align}

While process models may be preferable to simulation and linear regression approaches for predicting adoption outcomes of conservation interventions, they might not always be possible. In complex systems, these models can quickly require many parameters in order to accurately account for all moving parts and thus are prone to issues of identifiability in parameter estimation and may even become mathematically intractable [@laubmeier_ecological_2020]. Further, as has been demonstrated in coastal fisheries, these systems can exhibit mathematical chaos, where our inability to measure the parameters of interest at a fine enough resolution results in seemingly stochastic, unpredictable outcomes [@wilson_chaos_1994].

Coupled human and natural systems commonly exhibit dynamics with thresholds, resilience, and heterogeneity not present in social or natural systems in isolation and therefore are difficult to model analytically [@liu_complexity_2007]. In very complex cases such as these, we argue that agent-based simulations are the preferred approach as they allow for these features of complex systems which are difficult or even impossible to account for with process models such the SIRS or SIRS extension models presented above [@reyers_social-ecological_2018; @railsback_agent-based_2012; @wilensky_introduction_2015].
